Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-04 manuscript/response_to_reviews.Rmd

[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)

Only assessing models with AIC is not enough to prove sound relationships in multidimensional and complex data-models. LL ratio tests or deviance-anova tests can be used in the case of nested models with LMM or GLMM respectively. Along the text: dAIC is generally referred as $\Delta$AIC, please change accordingly.

[*issue #94*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/94)

**I have gone through the manuscript / tables and figures and updated where it said "dAICc" to be $\Delta$AICc. **

@@

*Ugghh, this doesn't even change anything--we'd still be analyzing the same 3 droughts! The question is whether to remove the growth reduction criteria. I personally think that analyzing 2 criteria is a strength, as our goal was to identify droughts that were both climatologically and biologically signficant. In any case, this is just an issue of changing the wording.*

*IM - I've updated the language in the manuscript, I think it's better.*

@@

Other minor comments:

- Line 113: correct ‘positively’. **done**

- Line 134: cambial growth also includes that of phloem; so better refer to ‘xylem growth’. **done**

- Line 161: how many species were estimated for height like this? Wouldn’t it be possible for authors to go to the field and perform some height estimations? This 
should be rather straightforward and the authors would increase accuracy across species in height estimation.
  - *IM I'm confused about how to answer this. There's no direct response to the first question, as the measurements I got (along with Atticus and previous measurements) were all lumped into the pot to make the regression equations, which we explain. Otherwise, I did go out and confirm measurements. I guess my only response (and question to you @teixeirak) is it worth explaining how disparate our data was and that's why we have different methods? Along with me going and getting new ones?*

- Lines 166-167: please explain further the hypothesis and why. *IM this is now line 170. I'm a little confused myself.*

- Line 222: what about climate in years before 1991? *IM I don't get this. At the beginning of the paragraph we say we looked at droughts 1950-2009...*

- Line 231: define AICc. **This is done (with added reference)**

- *IM I'm a bit confused by the next three comments.*
- Line 234-247: it is not clear what the authors did, please explain clearer this paragraph and the next and why you did two different (yet similar) analyses (single-variable and full model).

@@

- Line 255-256: but this was already an assumption in M-M (line 202-203). Please, see my comment below and be consistent along the text.

- Line 258: so this means that they are no sensitive to the drought? Which trees, any species in particular? Is this a consequence of stand dynamics shadowing the impact of drought events?

- Line 266-267: why only in 1966? I think this shows that this variable is not properly assessed. The older the date, the more accurate the estimation of crown position and height were.

*this is true of crown position, not height*
- *IM how do we want to officially respond to this?*

@@

- Line 275-280: those profiles were assumed a priori in the introduction and are common to any forest. They are indeed interesting to characterise the forest, but they do not add to individual tree variability. You could think to send Figure 2 to an appendix, in consequence, and use figures to show more specific results related to your individual tree analyses.

[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)

- Line 330-331: any evidence to support this? It looks rather speculative. As suggested above, does any species particularly show positive Rt or positive Rt are randomly distributed in your trees?

**IM I have done this (with arima ratio), see below. Assuming this should be SI plot?**
![Rt (Arima ratio) across species](tables_figures/rt_across_species.png)

- Line 331-333: not so sure about this. It can also mean that there is no causal relationship between Rt and covariates, mostly because radial growth Rt should not be used as an integral proxy of tree resistance to drought, such as it is often used now in the literature.
  - *IM Krista do you know more about this?*

- Line 334-339: indeed, assessment of these two variables is not conclusive, as suggested above (just data from 2018, from the past estimated and correlated). *IM does this need a response?*

- Line 348: non-drought. *IM do we have the original submitted draft? I'm not sure what this is referring to*

- Line 349: do authors really think that lower radial growth as from Rt on a specific year means that the tree is more vulnerable? I think this is over conclusive.
  - *IM well, in our original course of thinking about it as based on Lloret, yes, Rt is a measure of a tree's particular resistance, and it fluctuates from year to year. Maybe the reviewer would have preferred if we had said that lower radial growth shows the tree was more vulnerable *in that year*? And then we reiterate that the focal drought years all had low Rt?*
  - *That being said, we're now using arima ratio and all values of Rt are positive. Regardless, I don't see where we make the full claim the reviewer is referring to.*

- Line 400: I do not agree we can directly link Rt with tree vulnerability to drought; we might be currently overinterpreting these resilience indices extrapolating to physiological conclusions that do not necessary need to hold.
  - *this explains the previous comment*

Table 1: do ‘-‘ signs in the table correspond to ‘no’ (non-significant)? Please explain.
- *IM I added in a thing to the caption*

Table 3: please, add also the number of trees, it is more informative than the number of cores to check sample size in your study.
- *IM Krista, isn't the number of cores synonymous for us with the sample size for each species?*

Table 3: the authors could show some test to see if the traits presented are different among species.
- *IM Krista, do you think this is necessary? This would require extra stats to say something like "x traits did not differ significantly across species." I'm not sure how that result would affect our conclusions/writing.*

Table S1 and S2: correct R2. *IM I added the correct Latex notation but it's not rendering it when I knit the file...*

Table S3: please, explain what ‘rank’ means. *IM Krista, do you remember? Is this from the pointer year exercise? Or was this ranking the months in the context of each year's PDSI?*

Figure S3: I understand that authors only have height data from 2018 and height for the other dates looks trivial: indeed there will be an increase in height with an increase in DBH in time. The question is whether there was any stand dynamics involved in those height series that had to be included in the models.
- *IM seems like this comment is meant for the paper than for Figure S3. What do you think?*

### 2020-07-04 manuscript/climate_sensitivity_SI.Rmd
Added content based on Reviewer 1 comments.

bark <- read.csv("tables_figures/publication/tableS1_bark_regression.csv", stringsAsFactors = FALSE)
bark[,2] <- NULL
kable(bark, booktabs=TRUE, format = "latex", caption = "Table S1: Species-specific bark thickness regression equations") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

@@

heights <- read.csv("tables_figures/publication/tableS2_height_regression.csv", stringsAsFactors = FALSE)
heights[,2] <- NULL
kable(heights, booktabs=TRUE, format = "latex", caption = "Table S2: Species-specific height regression equations") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

\newpage
```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
tableS3 <- read.csv("tables_figures/publication/tables3_drought_stats.csv", stringsAsFactors = FALSE)
kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts") %>%
  pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
  pack_rows("", 5, 8, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 9, 12, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("other", 13, 16, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

>>>>

bark <- read.csv("tables_figures/publication/tableS1_bark_regression.csv", stringsAsFactors = FALSE)
bark[,2] <- NULL
colnames(bark) <- c("Species", "Equations", "$R^{2}$")
kable(bark, booktabs=TRUE, format = "latex", caption = "Table S1: Species-specific bark thickness regression equations") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

@@

heights <- read.csv("tables_figures/publication/tableS2_height_regression.csv", stringsAsFactors = FALSE)
heights[,2] <- NULL
colnames(heights) <- c("Species", "Equations", "$R^{2}$")
kable(heights, booktabs=TRUE, format = "latex", caption = "Table S2: Species-specific height regression equations") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

\newpage
```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
tableS3 <- read.csv("tables_figures/publication/tables3_drought_stats.csv", stringsAsFactors = FALSE)
kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts. Rank refers to ") %>%
  pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
  pack_rows("", 5, 8, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 9, 12, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("other", 13, 16, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

### 2020-07-04 manuscript/climate_sensitivity_draft.Rmd
Added content based on Reviewer 1 comments.

We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Fig. **S1**). Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017. In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead during annual mortality censuses [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies were published in Zenodo (DOI: 10.5281/zenodo.2649302) in association with @helcoski_growing_2019.

@@

Tree heights ($H$) were measured by several researchers for a variety of purposes between 2012 to 2019 (n=1,518 trees). Measurement methods included direct measurements using a collapsible measurement rod on small trees [@neon_national_2018] or a tape measure on recently fallen trees (this study); geometric calculations using clinometer and tape measure [@stovall_assessing_2018] or digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018]; and ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018]. Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H] ~ ln[DBH]$). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed by combining the height measurements across all species.

@@

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions and widespread reductions in tree growth, *i.e.*, droughts that substantially impacted the forest community. We identified three drought years: 1966, 1977, and 1999 (Figs. **1**, **S2**, Table S3). These were the three years with the lowest Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site. PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. These were also years with widespread tree growth reduction ("pointer years"), here defined as those where >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011]. Pointer years were identified using the pointRes package in R [@R-pointRes]. In addition to the focal drought years, 1991 also met this critera (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) but was excluded because it was not among the driest of the time period (Table S3). Rather, the severity of growth reduction could probably be explained in large part by defoliation by gypsy moths (*Lymantria dispar* L.), which was documented to have strongly impacted *Quercus* spp. in the area from approximately 1988 through 1995 [@twery_effects_1991].

@@

Analyses focused on testing the predictions presented in Table 1, with $Rt$ as the response variable. The general statistical model for hypothesis testing was a mixed effects model with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects. Mixed effects models were implemented in the lme4 package in R [@R-lme4]. We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit, implmented in the AICcmodavg package in R [@R-AICcmodavg].

Models were run for all drought years combined and for each drought year individually. In order to determine the relative importance of each predictor variable individually, we first implemented models with the variable in question as a fixed effect, along with drought year (for model with all drought years combined) and $ln[H]$ (included in null models because of it's substantial influence). Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 2 units relative to the corresponding null model lacking that variable (Table 4).

We then determined the best full models for predicting $Rt$ for each individual drought year and for all years combined. Candidate variables were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at dAICc $\ge$ 1.0) in at least one drought year (Table 4). We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models". When a variable appeared in all top models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction by the full models. If the variable appeared in only some of the models, we considered this partial support/rejection.

@@

The leaf hydraulic traits $PLA_{dry}$ and $\pi_{tlp}$ were linked to drought responses, whereas the other traits considered had insignficant and/or inconsistent correlations to $Rt$ (Tables 1,4,5). In the single-variable tests, $LMA$ and wood density were never significantly associated with $Rt$ (Table 4) and were excluded from the full models. In contrast, xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) during at least one of the three droughts (Table 4). Xylem porosity was not significant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5). $PLA_{dry}$ was a strong predictor for 1966 and all droughts combined, with consistently negative coefficients (Table 4). Similarly, $PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5). $\pi_{tlp}$ was not significant in any single-variable tests; however, coefficients were consistently negative (Table 4) and $\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).

>>>>

We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Fig. **S1**). Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017. In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead during annual mortality censuses [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies were published in Zenodo (DOI: 10.5281/zenodo.2649302) in association with @helcoski_growing_2019.

@@

Tree heights ($H$) were measured by several researchers for a variety of purposes between 2012 to 2019 (n=1,518 trees). Measurement methods included direct measurements using a collapsible measurement rod on small trees [@neon_national_2018] or a tape measure on recently fallen trees (this study); geometric calculations using clinometer and tape measure [@stovall_assessing_2018] or digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018]; and ground-based LiDAR [@stovall_terrestrial_2018]. Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other. Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018]. Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H] \sim ln[DBH]$). For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed by combining the height measurements across all species.

@@

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions and widespread reductions in tree growth, *i.e.*, droughts that substantially impacted the forest community. Using the pointRes package in R [@R-pointRes], we identified three drought years - 1966, 1977, and 1999 (Figs. **1**, **S2**, Table S3) - for two reasons. First, these were the three years with the lowest Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site. PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. Second, these were also years with widespread tree growth reduction, here defined as when >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years. This follows the drought resistance ($Rt$) metric of [@lloret_components_2011]. In addition to the focal drought years, 1991 also met this critera (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) but was excluded because it was not among the driest of the time period (Table S3). Rather, the severity of growth reduction could probably be explained in large part by defoliation by gypsy moths (*Lymantria dispar* L.), which was documented to have strongly impacted *Quercus* spp. in the area from approximately 1988 through 1995 [@twery_effects_1991].

@@

Analyses focused on testing the predictions presented in Table 1, with $Rt$ as the response variable. The general statistical model for hypothesis testing was a mixed effects model with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects. Mixed effects models were implemented in the lme4 package in R [@R-lme4]. We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implmented in the AICcmodavg package in R [@R-AICcmodavg]. AICc refers to a corrected version of AICc, and is best suited for small data sizes [see @brewer_relative_2016]

Models were run for all drought years combined and for each drought year individually. In order to determine the relative importance of each predictor variable individually, we first implemented models with the variable in question as a fixed effect, along with drought year (for model with all drought years combined) and $ln[H]$ (included in null models because of it's substantial influence). Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 1 unit relative to the corresponding null model lacking that variable (Table 4).

We then determined the best full models for predicting $Rt$ for each individual drought year and for all years combined. Candidate variables were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4). We compared models with all possible combinations of candidate variables and identified the full set of models within $\Delta$AICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models". When a variable appeared in all top models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction by the full models. If the variable appeared in only some of the models, we considered this partial support/rejection.

@@

The leaf hydraulic traits $PLA_{dry}$ and $\pi_{tlp}$ were linked to drought responses, whereas the other traits considered had insignficant and/or inconsistent correlations to $Rt$ (Tables 1,4,5). In the single-variable tests, $LMA$ and wood density were never significantly associated with $Rt$ (Table 4) and were excluded from the full models. In contrast, xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation ($\Delta$AICc > 1.0) during at least one of the three droughts (Table 4). Xylem porosity was not significant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5). $PLA_{dry}$ was a strong predictor for 1966 and all droughts combined, with consistently negative coefficients (Table 4). Similarly, $PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5). $\pi_{tlp}$ was not significant in any single-variable tests; however, coefficients were consistently negative (Table 4) and $\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).

### 2020-07-04 manuscript/tables&figures.Rmd
Added content based on Reviewer 1 comments.

```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/publication/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Table 1. Summary of hypotheses, corresponding specific predictions, and results. We count predictions as fully supported ('yes') when the response is signficant in single-variable tests (Table 4) and included in all top full models and as partially supported ('(yes)') or rejected ('(no)') when the direction of response consistently matched the prediction but the effect was not significant in all models.") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
  pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Larger, taller trees have lower Rt.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Trees with more exposed crowns have lower Rt.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Small trees (lower root volume) in drier microhabitats have lower Rt.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Species traits", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Species' traits--particularly leaf hydraulic traits--predict Rt.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1.2cm") %>%
  column_spec(6, width = "2.4cm") %>%
  kableExtra::landscape()
```

@@

\clearpage
```{r table 4, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
singlevar_table <- read.csv("tables_figures/publication/table4_singlevariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
kable(singlevar_table, booktabs = TRUE, caption = "Table 4. Single-variable tests of hypothesized drivers of drought resistance. Models including each variable were compared to corresponding null models. dAIC is the AICc of the null model minus that of the model including the variable (thus, dAICc>2 indicates that the variable significantly improves the model). Variable abbreviations are as in Table 2.", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
  # pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
  pack_rows("Species traits", 1, 6, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
  kable_styling()
```
*dAICc > 1: variable qualified for inclusion in full model

**dAICc > 2: statistically signficant, variable qualified for inclusion in full model

\clearpage
```{r table 5, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
kable(multivar_table, booktabs = TRUE, caption = "Table 5. Summary of top full models for each drought instance. Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," ", " ", " ", "crown position" = 4, 
                     "xylem porosity" = 2," ")) %>%
  pack_rows("", 1, 1, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 2, 2, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 3, 3, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 4, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  row_spec(1, bold = T) %>%
  row_spec(2, bold = T) %>%
  row_spec(3, bold = T) %>%
  row_spec(4, bold = T) %>%
  kable_styling()
```

>>>>

```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/publication/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Table 1. Summary of hypotheses, corresponding specific predictions, and results. We count predictions as fully supported ('yes') when the response is signficant in single-variable tests (Table 4) and included in all top full models and as partially supported ('(yes)') or rejected ('(no)') when the direction of response consistently matched the prediction but the effect was not significant in all models. A dash symbol indicates that the response was not significant in any of the top full models.") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
  pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Larger, taller trees have lower Rt.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Trees with more exposed crowns have lower Rt.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Small trees (lower root volume) in drier microhabitats have lower Rt.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Species traits", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Species' traits--particularly leaf hydraulic traits--predict Rt.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1.2cm") %>%
  column_spec(6, width = "2.4cm") %>%
  kableExtra::landscape()
```

@@

\clearpage
```{r table 4, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
singlevar_table <- read.csv("tables_figures/publication/table4_singlevariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
kable(singlevar_table, booktabs = TRUE, caption = "Table 4. Single-variable tests of hypothesized drivers of drought resistance. Models including each variable were compared to corresponding null models. Delta AICc is the AICc of the null model minus that of the model including the variable (thus, Delta AICc > 1 indicates that the variable significantly improves the model). Variable abbreviations are as in Table 2.", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
  # pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
  pack_rows("Species traits", 1, 6, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
  kable_styling()
```
*$\Delta$AICc > 1: variable qualified for inclusion in full model

\clearpage
```{r table 5, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
kable(multivar_table, booktabs = TRUE, caption = "Table 5. Summary of top full models for each drought instance. Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (Delta AICc<1) of the best model (bold).", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," ", " ", " ", "crown position" = 4, 
                     "xylem porosity" = 2," ")) %>%
  pack_rows("", 1, 1, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 2, 2, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 3, 3, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 4, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  row_spec(1, bold = T) %>%
  row_spec(2, bold = T) %>%
  row_spec(3, bold = T) %>%
  row_spec(4, bold = T) %>%
  kable_styling()
```
